---
title: "Chapter 10. Big Entropy and the Generalized Linear Model"
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output:
  github_document
bibliography: bib.bib
biblio-style: apalike
csl: apa.csl
link-citations: yes
---

```{r, echo = F, cache = F}
knitr::opts_chunk$set(fig.retina = 2.5)
knitr::opts_chunk$set(fig.align = "center")
options(width = 100)
```

# Big Entropy and the Generalized Linear Model

#### Rethinking: Bayesian updating is entropy maximization.

## Maximum entropy


```{r, message = F, warning = F}
library(tidyverse)
library(ghibli)
```

#### Rethinking: What good is intuition?

### Gaussian.

### Binomial.

The binomial likelihood entails

> counting the numbers of ways that a given observation could arise, according to our assumptions. The resulting distribution is known as the **binomial distribution**. If only two things can happen (blue or white marble, for example), and there's a constant chance $p$ of each across $n$ trials, then the probability of observing $y$ events of type 1 and $n - y$ events of type 2 is: 
>
> $$\text{Pr} (y | n, p) = \frac{n!}{y! (n - y)!} p^y (1 - p)^{n - y}$$
>
> It may help to note that the fraction with the factorials is just saying how many different ordered sequences of $n$ outcomes have a count of $y$. (p. 307, **emphasis** in the original)

For me, that last sentence made more sense when I walked it out in an example. To do so, let's wrap that fraction of factorials into a function.

```{r}
count_ways <- function(n, y) {
  # n = the total number of trials (i.e., the number of rows in your vector)
  # y = the total number of 1s (i.e., successes) in your vector
  (factorial(n) / (factorial(y) * factorial(n - y)))
}
```

Now consider three sequences:

* 0, 0, 0, 0 (i.e., $n = 4$ and $y = 0$)
* 1, 0, 0, 0 (i.e., $n = 4$ and $y = 1$)
* 1, 1, 0, 0 (i.e., $n = 4$ and $y = 2$)

We can organize that information in a little tibble and then demo our `count_ways()` function.

```{r}
tibble(sequence = 1:3,
       n        = 4,
       y        = c(0, 1, 2)) %>% 
  mutate(n_ways = count_ways(n = n, y = y))
```

Here's the pre-Figure 10.3 data McElreath presented on page 308.

```{r}
# data
d <-
  tibble(distribution = letters[1:4],
         ww = c(1/4, 2/6, 1/6, 1/8),
         bw = c(1/4, 1/6, 2/6, 4/8),
         wb = c(1/4, 1/6, 2/6, 2/8),
         bb = c(1/4, 2/6, 1/6, 1/8))

# table
d %>% 
  mutate_if(is.numeric, ~MASS::fractions(.) %>% as.character()) %>% 
  flextable::flextable()
```

Those data take just a tiny bit of wrangling before they're ready to plot in our version of Figure 10.3.

```{r, fig.width = 4, fig.height = 3.5}
d <- 
  d %>% 
  pivot_longer(-distribution,
               names_to = "sequence", 
               values_to = "probability") %>% 
  mutate(sequence = factor(sequence, levels = c("ww", "bw", "wb", "bb")))

d  %>% 
  ggplot(aes(x = sequence, y = probability, group = 1)) +
  geom_point(size = 2, color = ghibli_palette("PonyoMedium")[4]) +
  geom_line(color = ghibli_palette("PonyoMedium")[5]) +
  labs(x = NULL, y = NULL) +
  coord_cartesian(ylim = 0:1) +
  theme(axis.ticks.x = element_blank(),
        panel.background = element_rect(fill = ghibli_palette("PonyoMedium")[2]),
        panel.grid = element_blank(),
        strip.background = element_rect(fill = ghibli_palette("PonyoMedium")[6])) +
  facet_wrap(~ distribution)
```

```{r, echo = F, eval = F}
rm(d, alpha_per_beta, count_ways, p, sim_p, n_rep, p1, p2, ranked_d, subset_d, a, length_out, alpha, beta, lines, b10.1, post)
```

```{r, echo = F, message = F, warning = F, results = "hide", eval = F}
ggplot2::theme_set(ggplot2::theme_grey())
bayesplot::color_scheme_set("blue")
# pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

