---
title: "Section 13.5.2.1: Bonus: Let's use `fitted()` this time."
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output:
  github_document
bibliography: bib.bib
biblio-style: apalike
csl: apa.csl
link-citations: yes
---

```{r, echo = F, cache = F}
knitr::opts_chunk$set(fig.retina = 2.5)
knitr::opts_chunk$set(fig.align = "center")
options(width = 100)
```

```{r, message = F, warning = F}
# load the packages that would have already been loaded
library(brms)
library(tidyverse)
library(ggthemes) 
library(tidybayes)
library(bayesplot)
library(posterior)
library(patchwork)

# change the default
theme_set(theme_gray() + theme_fivethirtyeight())

# load and wrangle the data
data(chimpanzees, package = "rethinking")
d <- chimpanzees
rm(chimpanzees)

d <-
  d %>% 
  mutate(actor     = factor(actor),
         block     = factor(block),
         treatment = factor(1 + prosoc_left + 2 * condition))

# load the model
b13.4 <- 
  brm(data = d, 
      family = binomial,
      bf(pulled_left | trials(1) ~ a + b,
         a ~ 1 + (1 | actor) + (1 | block), 
         b ~ 0 + treatment,
         nl = TRUE),
      prior = c(prior(normal(0, 0.5), nlpar = b),
                prior(normal(0, 1.5), class = b, coef = Intercept, nlpar = a),
                prior(exponential(1), class = sd, group = actor, nlpar = a),
                prior(exponential(1), class = sd, group = block, nlpar = a)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 13,
      file = "/Users/solomonkurz/Dropbox/Recoding Statistical Rethinking 2nd ed/fits/b13.04")

# define the labels
labels <- c("R/N", "L/N", "R/P", "L/P")
```

# Models With Memory

## Example: Multilevel tadpoles

#### Overthinking: Prior for variance components.

## Varying effects and the underfitting/overfitting trade-off

### The model.

### Sumulate survivors.

### Compute the no-pooling estimates.

### Compute the partial-pooling estimates.

#### Overthinking: Repeating the pond simulation.

## More than one type of cluster

#### Rethinking: Cross-classification and hierarchy.

### Multilevel chimpanzees.

### Even more clusters.

## Divergent transitions and non-centered priors

### The Devil's Funnel.

### Non-centered chimpanzees.

## Multilevel posterior predictions

### Posterior prediction for same clusters.

### Posterior prediction for new clusters. 

#### Bonus: Let's use `fitted()` this time.

We just made those plots using various wrangled versions of `post`, the data frame returned by `posterior_samples(b.13.4)`. If you followed along closely, part of what made that a great exercise is that it forced you to consider what the various vectors in `post` meant with respect to the model formula. But it's also handy to see how to do that from a different perspective. So in this section, we'll repeat that process by relying on the `fitted()` function, instead. We'll go in the same order, starting with the average actor.

```{r, warning = F, message = F}
nd <- distinct(d, treatment)

(
  f <-
  fitted(b13.4,
         newdata = nd,
         re_formula = NA,
         probs = c(.1, .9)) %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  mutate(treatment = factor(treatment, labels = labels))
)
```

You should notice a few things. Since `b13.4` is a cross-classified multilevel model, it had three predictors: `treatment`, `block`, and `actor`. However, our `nd` data only included the first of those three. The reason `fitted()` permitted that was because we set `re_formula = NA`. When you do that, you tell `fitted()` to ignore group-level effects (i.e., focus only on the fixed effects). This was our `fitted()` version of ignoring the `r_` vectors returned by `posterior_samples()`. Here's the plot.

```{r, fig.width = 2.5, fig.height = 3.25}
p4 <-
  f %>%
  ggplot(aes(x = treatment, y = Estimate, group = 1)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), fill = "blue") +
  geom_line(color = "orange1") +
  ggtitle("Average actor") +
  coord_cartesian(ylim = 0:1) +
  theme(plot.title = element_text(size = 14, hjust = .5))

p4
```

For marginal of actor, we can continue using the same `nd` data. This time we'll be sticking with the default `re_formula` setting, which will accommodate the multilevel nature of the model. However, we'll also be adding `allow_new_levels = T` and `sample_new_levels = "gaussian"`. The former will allow us to marginalize across the specific actors and blocks in our data and the latter will instruct `fitted()` to use the multivariate normal distribution implied by the random effects. It'll make more sense why I say *multivariate* normal by the end of the [next chapter][Adventures in Covariance]. For now, just go with it.

```{r, warning = F, message = F}
(
  f <-
  fitted(b13.4,
         newdata = nd,
         probs = c(.1, .9),
         allow_new_levels = T,
         sample_new_levels = "gaussian") %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  mutate(treatment = factor(treatment, labels = labels))
  )
```

Here's our `fitted()`-based marginal of `actor` plot.

```{r, fig.width = 2.5, fig.height = 3.25}
p5 <-
  f %>%
  ggplot(aes(x = treatment, y = Estimate, group = 1)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), fill = "blue") +
  geom_line(color = "orange1") +
  ggtitle("Marginal of actor") +
  coord_cartesian(ylim = 0:1) +
  theme(plot.title = element_text(size = 14, hjust = .5))

p5
```

We'll have to amend our workflow a bit to make a `fitted()` version of the third panel. First we redefine our `nd` data and execute the `fitted()` code.

```{r}
# how many simulated chimps would you like?
n_chimps <- 100

nd <- 
  distinct(d, treatment) %>% 
  # define 100 new actors
  expand(actor = str_c("new", 1:n_chimps),
         treatment) %>% 
  # this adds a row number, which will come in handy, later
  mutate(row = 1:n())

# fitted
set.seed(13)

f <-
  fitted(b13.4,
         newdata = nd,
         allow_new_levels = T,
         sample_new_levels = "gaussian",
         summary = F,
         nsamples = n_chimps)
```

Our `f` object will need a lot of wrangling. Before I walk out the wrangling steps, we should reiterate what McElreath originally did in the text (pp. 429--430). He based the new actors on the deviation scores from `post$sigma_a`. That was the first working line in his **R** code 13.38. In the remaining lines in that code block, he used the model formula to compute the actor-level trajectories. Then in his plot code in **R** code 13.39, he just used the first 100 rows from that output. 

In our `fitted()` code, above, we saved a little time and computer memory by setting `nsamples = n_chimps`, which equaled 100. That's functionally the same as when McElreath used the first 100 posterior draws in the plot. A difficulty for us is the way `brms::fitted()` returns the output, the 100 new levels of `actor` and the four levels of `treatment` are confounded in the 400 columns. In the code block, below, the `data.frame()` through `left_join()` lines are meant to disentangle those two. After that, we'll make an `actor_number` variable, which which we'll filter the data such that the first row returned by `fitted()` is only assigned to the new actor #1, the second row is only assigned to the new actor #2, and so on. The result is that we have 100 new simulated actors, each of which corresponds to a different iteration of the posterior draws from the fixed effects[^1].

```{r, fig.width = 2.5, fig.height = 3.25}
p6 <-
  f %>%
  data.frame() %>% 
  # name the columns by the `row` values in `nd`
  set_names(pull(nd, row)) %>% 
  # add an iteration index
  mutate(iter = 1:n()) %>% 
  # make it long
  pivot_longer(-iter, names_to = "row") %>% 
  mutate(row = as.double(row)) %>% 
  # add the new data
  left_join(nd, by = "row") %>% 
  # extract the numbers from the names of the new actors
  mutate(actor_number = str_extract(actor, "\\d+") %>% as.double()) %>% 
  # only keep the posterior iterations that match the `actor_number` values
  filter(actor_number == iter) %>% 
  # add the `treatment` labels
  mutate(treatment = factor(treatment, labels = labels)) %>% 
  
  # plot!
  ggplot(aes(x = treatment, y = value, group = actor)) +
  geom_line(alpha = 1/2, color = "blue") +
  ggtitle("100 simulated actors") +
  theme(plot.title = element_text(size = 14, hjust = .5))

p6
```

Here they are altogether.

```{r, fig.width = 7.5, fig.height = 3.25}
p4 | p5 | p6
```

## Session info

```{r}
sessionInfo()
```

## Endnote

[^1]: The `fitted()` version of the code for the third panel is cumbersome. Indeed, this in one of those cases where it seems more straightforward to work directly with the `posterior_samples()` output, rather than with `fitted()`. The workflow in this section from previous editions of this ebook was more streamlined and superficially seemed to work. However, fellow researcher [Ladislas Nalborczyk](https://twitter.com/lnalborczyk) kindly pointed out I was taking 100 draws from one new simulated `actor`, rather than one simulated draw from 100 new levels of `actor`. To my knowledge, if you want 100 new levels of `actor` AND want each one to be from a different posterior iteration, you'll need a lot of post-processing code when working with `fitted()`.

